---
title: "Celegans_filtering"
author: "Laura Biggins"
date: "2023-05-16"
output: html_document
---

---
title: "C elegans scRNA"
author: "Laura Biggins"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(limma)
library(tidyverse)
library(DT)
```

## Finding genes that are upregulated in certain clusters

## Identified genes

We've got sets of marker genes that were identified using the Seurat function `FindMarkers()`.
From these, we want to do some more filtering to find genes that are consistently 
highly expressed across each of the time points.

```{r}
# for displaying tables (in markdown doc)
show_up_genes <- function(markers){
	up_genes <- markers %>%
  	filter(avg_log2FC > 0) %>%
  	arrange(desc(avg_log2FC)) 

	DT::datatable(up_genes)
}

# for actually returning the tbl
get_up_genes <- function(markers){
	up_genes <- markers %>%
  	filter(avg_log2FC > 0) %>%
  	arrange(desc(avg_log2FC)) 
}

```


### intestine

```{r}
load("data/intestine_markers.rds")
int <- get_up_genes(intestine_markers)
```


```{r, eval=FALSE}
# Extracting the counts from the matrix
# only running this once as it uses a lot of memory
load("data/clr_norm_scaled_worms.rds")
worm_counts <- worms@assays$RNA@counts
save(worm_counts, file = "data/worm_counts.rds")
rm(worms) # it's big so we don't want to keep it in memory if we don't need it
```

```{r}
load("data/worm_counts.rds")
```

Extract the counts for the genes that were up in the intestine clusters

```{r}
int_mat <- worm_counts[rownames(int),]

# I'm assuming d5_1 and d5_2 are reps 1 and 2.
intestine_tbl <- as_tibble(int_mat[,], rownames="gene") %>%
	pivot_longer(cols=-gene) %>%
	separate_wider_delim(cols=name, delim="_", names = c(NA, "day", "rep")) %>%
  mutate(day = factor(day, levels = c("d1","d3","d5","d8","d11","d15"))) 

							
intestine_tbl %>%
  count(day, rep) %>%
	ggplot(aes(x=day, y=n, fill=rep)) +
	geom_col() +
	scale_fill_brewer(palette="Set1") +
	ylab("Read count")
```

There isn't really any data for d15 so we need to bear this in mind.

```{r}
summarised_inc0 <- intestine_tbl %>%
	mutate(day = factor(day, levels = c("d1","d3","d5","d8","d11","d15"))) %>%
	#group_by(gene, day, rep) %>%
	group_by(gene, day) %>%
	summarise(
	  mean = mean(value),
	  median = median(value),
	  sd = sd(value)
	) %>%
	ungroup()

knitr::kable(head(summarised_inc0))
```

All the medians are 0 as there are so many with 0 counts, which isn't surprising for single cell data.

We'll do another summary but removing all the 0 counts.
We can filter on a minimum count value later.

```{r}
summarised_no0 <- intestine_tbl %>%
	filter(value > 0) %>%
	mutate(day = factor(day, levels = c("d1","d3","d5","d8","d11","d15"))) %>%
	#group_by(gene, day, rep) %>%
	group_by(gene, day) %>%
	summarise(
	  mean = mean(value),
	  median = median(value),
	  sd = sd(value),
	  count = n()
	) %>%
	ungroup()


summarised_no0 %>%
	ggplot(aes(x=day, y=mean, group=gene)) +
  geom_line()


```

This looks a bit odd - there's something with a very high count for d15 - we probably want to get rid of that.

```{r}
summarised_no0 %>%
	arrange(desc(mean)) %>%
	head() %>%
	knitr::kable()
#intestine_tbl
```
    
It's dod-6 that has the really high mean values, that's going to account for most of the counts in d15.
I'll remove it so that we can see the other genes more clearly.

```{r}
summarised_no0 <- summarised_no0 %>%
	filter(gene != "dod-6")

summarised_no0 %>%
	ggplot(aes(x=day, y=mean, group=gene)) +
  geom_line()
```


## Counts

```{r}
summarised_no0 %>%
	ggplot(aes(x=day, y = log2(count))) +
	geom_violin() +
	geom_hline(yintercept = 6, colour = "red2", lty="dashed")
```

We'll filter where count > 64. 


## Mean values

```{r}
summarised_no0 %>%
	ggplot(aes(x=day, y = log2(mean))) +
	geom_violin()
```

## Median values

```{r}
summarised_no0 %>%
	ggplot(aes(x=day, y = log2(median))) +
	geom_violin() +
	geom_jitter(height=0, alpha=0.5)
```

There are a lot of median values of 1 (showing as 0 on the plot as log transformed), let's try filtering where median > 1.

We'll remove d15 from the filtering, then select where all counts are > 64

```{r}
high_counts_all_days <- summarised_no0 %>%
	filter(day != "d15") %>%
	filter(count > 64) %>%
	filter(median > 1) %>%
	group_by(gene) %>%
	summarise(count_days = n()) %>%
	filter(count_days == 5) %>%
	pull(gene)
```

We've got `r nrow(high_counts_all_days)` genes with high overall counts and medians of > 1 in d1-d11. 

We don't need to redo the summarisation, we'll just filter it for the genes we've selected.

```{r}
summarised_no0 %>%
	filter(gene %in% high_counts_all_days) %>%
	DT::datatable() %>%
	DT::formatRound(columns=c('mean', 'sd'), digits=3)
```


```{r}
summarised_no0 %>%
	filter(gene %in% high_counts_all_days) %>%
	select(-median, -count) %>%
	pivot_wider(
		names_from = day, 
		values_from = c(mean, sd)
	) %>%
	DT::datatable() %>%
	DT::formatRound(columns=2:13, digits=2)
```


```{r}
summarised_no0 %>%
	filter(gene %in% high_counts_all_days) %>%
	ggplot(aes(x=day, y=mean, group=gene, col=gene)) +
  geom_line()

summarised_no0 %>%
	filter(gene %in% high_counts_all_days) %>%
	ggplot(aes(x=day, y=median, group=gene, col=gene)) +
  geom_line()
```


The same plots but including all the other data
```{r}
subset <- summarised_no0 %>%
	filter(gene %in% high_counts_all_days)

summarised_no0 %>%
	ggplot(aes(x=day, y=mean, group=gene)) +
	geom_line(alpha = 0.5) +
  geom_line(data=subset, aes(col=gene))
```

```{r}
summarised_no0 %>%
	ggplot(aes(x=day, y=median, group=gene)) +
	geom_line(alpha = 0.5) +
  geom_line(data=subset, aes(col=gene))
```

There are a lot of lines on top of each other as most of the median values are 1 or 2. 



```{r, eval = FALSE, echo=FALSE}	
# check whether the values are different in this normalised data vs the original data

# go to line 220	


summarised_gene <- summarised_no0 %>%
	filter(count > 16) %>%
	group_by(gene) %>%
	summarise(
		mean_of_mean = mean(mean),
		mean_of_median = mean(median),
		mean_of_sd = mean(sd)
	) 

#plot the mean vs sd for these genes
summarised_gene %>%
	ggplot(aes(x=mean, y=mean_of_sd)) +
	geom_point()

summarised_gene %>%
	ggplot(aes(x=mean_of_median, y=mean_of_sd)) +
	geom_point()


top_genes <- summarised_gene %>%
	arrange(desc(mean_of_median)) %>%
	slice(1:10) %>%
	pull(gene)
```



### body wall muscle

```{r}
load("data/body_wall_muscle_markers.rds")
show_up_genes(body_wall_muscle_markers)
```


### pharynx

```{r}
load("data/pharynx_markers.rds")
show_up_genes(pharynx_markers)
```


### neurons

```{r}
load("data/neuron_markers.rds")
show_up_genes(neuron_markers)
```