---
title: "C elegans scRNA"
author: "Laura Biggins"
date: "2023-04-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(limma)
library(tidyverse)
library(DT)
```

## Finding genes that are upregulated in certain clusters


```{r, eval = FALSE}
#load("data/clr_norm_scaled_worms.rds")
load("data/clr_norm_scaled_worms_lax.rds")

meta <- worms@meta.data
cluster_names <- as.character(meta$annotate_name)
```


Using the FindMarkers Seurat function

### intestine

```{r, eval = FALSE}
of_interest <- grepl("^3_[1245]", cluster_names) # intestine

x <- meta[of_interest , ]
y <- meta[!of_interest , ]

#intestine_markers0 <- FindMarkers(worms@assays$RNA, cells.1 = rownames(x), cells.2 = rownames(y))
intestine_markers <- FindMarkers(
	worms@assays$RNA, 
	cells.1 = rownames(x), 
	cells.2 = rownames(y),
	test.use = "roc", 
	only.pos = TRUE
)

save(intestine_markers, file = "data/intestine_markers_roc.rds")
```

### body wall muscle

```{r, eval=FALSE}
name_pattern <- "^4_[012]|^1_1"

of_interest <- grepl(name_pattern, cluster_names)
# check
unique(cluster_names[of_interest])

body_wall_muscle_markers <- FindMarkers(
  worms@assays$RNA, 
  cells.1 = rownames(meta)[of_interest], 
  cells.2 = rownames(meta)[!of_interest],
  test.use = "roc", 
	only.pos = TRUE
)

save(body_wall_muscle_markers, file = "data/body_wall_muscle_markers_roc.rds")
```

### pharynx

```{r, eval=FALSE}
name_pattern <- "^26_[012]|^24_1|^41_0|^78_0"

of_interest <- grepl(name_pattern, cluster_names)

unique(cluster_names[of_interest]) # check

pharynx_markers <- FindMarkers(
  worms@assays$RNA, 
  cells.1 = rownames(meta)[of_interest], 
  cells.2 = rownames(meta)[!of_interest],
  test.use = "roc", 
	only.pos = TRUE
)

save(pharynx_markers, file = "data/pharynx_markers_roc.rds")
```

### neurons

```{r, eval = FALSE}
name_pattern <- "^6_[0123]|^10_0|^14_[012]|^1[79]_0|^20_[012]|^22_0|^25_[01]|^2[89]_0|^32_0|^35[01234]|^36_[023]|^39_[0123]|^44_[01]|^46_0|^47_[01]|^4[89]_0|^5[01245]_0|^51_1|^6[1235689]_0|^70_[01]|^7[1234567]_0|^8[0134567]_0|^9[01234789]_0|^10[01346789]_0|^11[1234578]_0|^12[0123789]_0|^13[012345678]_0|^14[0245]_0"

of_interest <- grepl(name_pattern, cluster_names)

unique(cluster_names[of_interest]) # check

neuron_markers <- FindMarkers(
  worms@assays$RNA, 
  cells.1 = rownames(meta)[of_interest], 
  cells.2 = rownames(meta)[!of_interest],
  test.use = "roc", 
	only.pos = TRUE
)

save(neuron_markers, file = "data/neuron_markers_roc.rds")
```


## Identified genes

```{r}
show_up_genes <- function(markers){
	up_genes <- markers %>%
  	filter(avg_log2FC > 0) %>%
  	arrange(desc(avg_log2FC)) 

	DT::datatable(up_genes)
}
```


### intestine

```{r}
load("data/intestine_markers_roc.rds")
DT::datatable(intestine_markers)
#show_up_genes(intestine_markers)
```

Looking at vit-6 on the calcicolabs interface, it does seem to be a good marker, but cluster 3_5 has very little data and so does not show as being expressed at the later time points.

### body wall muscle

```{r}
load("data/body_wall_muscle_markers_roc.rds")
DT::datatable(body_wall_muscle_markers)
#show_up_genes(body_wall_muscle_markers)
```


### pharynx

```{r}
load("data/pharynx_markers_roc.rds")
DT::datatable(pharynx_markers)
#show_up_genes(pharynx_markers)
```


### neurons

```{r}
load("data/neuron_markers_roc.rds")
DT::datatable(neuron_markers)
#show_up_genes(neuron_markers)
```














```{r, echo = FALSE, eval=FALSE}
rownames(up_genes)

all_cells <- as.matrix(worms@assays$RNA@counts[rownames(up_genes), ])

#one_gene  <- all_cells[1,]

tbl_selected <- as_tibble(all_cells, rownames = "gene") %>%
	pivot_longer(cols = -gene, values_to = "count", names_to = "cell") %>%
  mutate(type = ifelse(cell %in% rownames(x), "marker", "other"))

tbl_intestine <- tbl_selected %>%
  filter(type == "marker")


tbl_selected %>%
	ggplot(aes(x = gene, y = log2(count))) +
	geom_boxplot(outlier.shape = NA, alpha = 0.5) +
	geom_jitter(data = tbl_intestine, size = 1, colour = "blue", alpha = 0.5, width = 0.2, height = 0) +
	coord_flip()

tbl_selected %>%
	ggplot(aes(x = gene, y = log2(count), colour = type)) +
	geom_boxplot(outlier.shape = NA, alpha = 0.5)

VlnPlot(worms,features="dct-16")

```







