---
title: "Tables of marker genes"
date: "2023-06-14"
output:
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE)
first_run <- FALSE
library(tidyverse)

load("data/intestine_table.rds")
load("data/bw_table.rds")
load("data/pharynx_table.rds")
load("data/neuron_table.rds")
load("data/cluster_names.rds")
```

I tried rerunning the ROC analysis, this time including the clusters in white (from the Excel sheet) for the tissues of interest. Overall, this didn't seem to help the analysis, so for the main tables, I went back to running the ROC analysis on the clusters highlighted in green. The [results from the extended sets](#extended_clusters) can be seen further down the page. 
The clusters in white for the tissues of interest are named `intestine_other`, `bw_other` and `pharynx_other`.

```{r functions, eval=TRUE}
get_timepoint_non_zeros <- function(marker_matrix, clusters_of_interest){

	# I'm assuming d5_1 and d5_2 are reps 1 and 2.
	as_tibble(marker_matrix[,], rownames="gene") %>%
		pivot_longer(cols=-gene) %>%
		filter(name %in% clusters_of_interest) %>%
		separate_wider_delim(cols=name, delim="_", names = c(NA, "day", "rep")) %>%
	  mutate(day = factor(day, levels = c("d1","d3","d5","d8","d11","d15"))) %>%
		mutate(non_zero = value>0) %>%
		group_by(gene, day) %>%
		summarise(
		  n = n(),
		  non_zero_count = sum(non_zero)
		) %>%
		ungroup() %>%
		mutate(pct = 100*(non_zero_count/n)) %>%
		select(-n, -non_zero_count) 
}
	
add_min_column <- function(timepoint_non_zeros){
	timepoint_non_zeros %>%
		group_by(gene) %>%
		summarise(min_day = min(pct)) %>%
		ungroup()
}
	
timepoint_wrapper <- function(count_table, marker_genes, clusters_of_interest){
	
	marker_matrix <- as.matrix(count_table[rownames(marker_genes),])
	non_zeros <- get_timepoint_non_zeros(marker_matrix, clusters_of_interest)
	timepoint_mins <- add_min_column(non_zeros)
	
	non_zeros <- non_zeros %>%
		pivot_wider(names_from = day, values_from = pct, names_prefix = "non0_") %>%
		left_join(timepoint_mins)
	
	marker_genes %>%
		tibble::rownames_to_column(var = "gene") %>%
		full_join(non_zeros) 
}

formatDT <- function(tbl){
	tbl %>%
		relocate(gene, pct.1, pct.2, myAUC, power, min_day) %>%
		relocate(avg_diff, avg_log2FC, .after=last_col()) %>%
		DT::datatable(rownames = FALSE, filter = "top") %>%
		DT::formatRound(columns=2:ncol(tbl), digits=3)
}

formatDT_no_relocate <- function(tbl){
	tbl %>%
		DT::datatable(rownames = FALSE, filter = "top", options = list(scrollX = TRUE)) %>%
		DT::formatRound(columns=2:ncol(tbl), digits=3)
}

```

```{r, eval = FALSE}
load("data/clr_norm_scaled_worms_lax.rds")
worm_counts_qc_filt_lax <- worms@assays$RNA@counts
meta <- worms@meta.data
save(meta, file = "data/worm_meta.rds")
save(worm_counts_qc_filt_lax, file = "data/worm_counts_qc_filt_lax.rds")
```

```{r, eval=first_run}
# process the data once to get the timepoint info
load("data/worm_counts_qc_filt_lax.rds")
load("data/intestine_markers_roc.rds")
load("data/body_wall_muscle_markers_roc.rds")
load("data/pharynx_markers_roc.rds")
load("data/neuron_markers_roc.rds")
load("data/cluster_names.rds")
load("data/worm_meta.rds")

of_interest <- grepl("^3_[1245]", cluster_names) # intestine
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
intestine_table <- timepoint_wrapper(worm_counts_qc_filt_lax, intestine_markers, x)

name_pattern <- "^4_[012]|^1_1"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
bw_table <- timepoint_wrapper(worm_counts_qc_filt_lax, body_wall_muscle_markers, x)

name_pattern <- "^26_[012]|^24_1|^41_0|^78_0"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
pharynx_table <- timepoint_wrapper(worm_counts_qc_filt_lax, pharynx_markers, x)

name_pattern <- "^6_[0123]|^10_0|^14_[012]|^1[79]_0|^20_[012]|^22_0|^25_[01]|^2[89]_0|^32_0|^35_[01234]|^36_[023]|^39_[0123]|^44_[01]|^46_0|^47_[01]|^4[89]_0|^5[01245]_0|^51_1|^6[1235689]_0|^70_[01]|^7[1234567]_0|^8[0134567]_0|^9[01234789]_0|^10[01346789]_0|^11[1234578]_0|^12[0123789]_0|^13[012345678]_0|^14[0245]_0"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
neuron_table <- timepoint_wrapper(worm_counts_qc_filt_lax, neuron_markers, x)

save(intestine_table, file = "data/intestine_table.rds")
save(bw_table, file = "data/bw_table.rds")
save(pharynx_table, file = "data/pharynx_table.rds")
save(neuron_table, file = "data/neuron_table.rds")
```

# Classification of clusters

We want to include a breakdown of expression across all tissues, so we're using all the annotated clusters and grouping them more generally into tissues.

```{r, echo =TRUE}
intestine_other <- "^30_[04]|^1_2|^38_0"
bw_other <- "^7_[01]|^143_0"
pharynx_other <- "^1_0|^9_[012]|^15_0|^24_0|^41_1"
epithelium <- "^11_[12]|^13_1|^27_0|^41_2|^57_0|^68_1|^146_0"
glia <- "^5_[01]|^16_0|^21_0|^30_5|^34_[01]|^37_0|^45_0|^79_[02]|^88_0|^141_0"
hypodermis <- "^8_0|^30_[12]|^33_[01]|^2_0|^64_0"
muscle_excl_bw_pharyngeal <- "^7_[234]|^38_[123]|^102_0"
gonad <- "^18_0|^31_0|^43_0"
germline <- "^3_[03]|^57_1"
mesoderm <- "^23_0|^110_1|^119_0"
coelomocyte <- "^110_0|^12_0"
gland <- "^89_0|^30_3"
seam <- "^0_0|^11_3"
valve <- "^60_0"
vulva <- "^11_0"
excretory <- "^79_1"
endocrine <- "^40_0"
other <- "^13_0|^21_1|^42_0|^82_[01]"
other_neurons <- "^5[369]_0|^58_[01]|^67_0|^9[56]_0|^105_0|^116_0|^12[456]_0|^139_0"

intestine <- "^3_[1245]"
body_wall <- "^4_[012]|^1_1"
pharynx <- "^26_[012]|^24_1|^41_0|^78_0"
neurons <- "^6_[0123]|^10_0|^14_[012]|^1[79]_0|^20_[012]|^22_0|^25_[01]|^2[89]_0|^32_0|^35_[01234]|^36_[023]|^39_[0123]|^44_[01]|^46_0|^47_[01]|^4[89]_0|^5[01245]_0|^51_1|^6[1235689]_0|^70_[01]|^7[1234567]_0|^8[0134567]_0|^9[01234789]_0|^10[01346789]_0|^11[1234578]_0|^12[0123789]_0|^13[012345678]_0|^14[0245]_0"
```


```{r}
load("data/worm_counts_qc_filt_lax.rds")
load("data/worm_meta.rds")

all_clusters_categorised <- tibble::tibble(
	cluster_name = meta$annotate_name) %>%
	add_column("cell_name" = colnames(worm_counts_qc_filt_lax)) %>%
	mutate(
		category = case_when(
			grepl(intestine_other, cluster_names) ~ "intestine_other",
			grepl(bw_other, cluster_names) ~ "bw_other",
			grepl(pharynx_other, cluster_names) ~ "pharynx_other",
			grepl(epithelium, cluster_names) ~ "epithelium",
			grepl(glia, cluster_names) ~ "glia",
			grepl(hypodermis, cluster_names) ~ "hypodermis",
			grepl(muscle_excl_bw_pharyngeal, cluster_names) ~ "muscle_excl_bw_pharyngeal",
			grepl(gonad, cluster_names) ~ "gonad",
			grepl(germline, cluster_names) ~ "germline",
			grepl(mesoderm, cluster_names) ~ "mesoderm",
			grepl(coelomocyte, cluster_names) ~ "coelomocyte",
			grepl(gland, cluster_names) ~ "gland",
			grepl(seam, cluster_names) ~ "seam",
			grepl(valve, cluster_names) ~ "valve",
			grepl(vulva, cluster_names) ~ "vulva",
			grepl(excretory, cluster_names) ~ "excretory",
			grepl(endocrine, cluster_names) ~ "endocrine",
			grepl(other, cluster_names) ~ "other",
			grepl(other_neurons, cluster_names) ~ "other_neurons",
			grepl(intestine, cluster_names) ~ "intestine",
			grepl(body_wall, cluster_names) ~ "body_wall",
			grepl(pharynx, cluster_names) ~ "pharynx",
			grepl(neurons, cluster_names) ~ "neurons", 
			.default = "not_matched"
		)
	)

knitr::kable(head(all_clusters_categorised))
```

All of the `r nrow(all_clusters_categorised)` cells in this dataset seem to have been identified, the table above just shows the first few lines.
 
## Number of cells in each cluster

```{r}
all_clusters_categorised %>%
	group_by(category) %>%
	count(name="number of cells") %>%
	arrange(desc(`number of cells`)) %>%
	DT::datatable(rownames = FALSE)
```



```{r}
load("data/intestine_markers_roc.rds")
load("data/body_wall_muscle_markers_roc.rds")
load("data/pharynx_markers_roc.rds")
load("data/neuron_markers_roc.rds")
load("data/cluster_names.rds")

get_all_tissues_non0 <- function(marker_matrix){
	as_tibble(marker_matrix[,], rownames="gene") %>%
	pivot_longer(cols=-gene, names_to = "cell_name") %>%
	left_join(all_clusters_categorised) %>%
	mutate(non_zero = value>0) %>%
	group_by(gene, category) %>%
	summarise(
	  n = n(),
	  non_zero_count = sum(non_zero)
	) %>%
	ungroup() %>%
	mutate(pct = 100*(non_zero_count/n)) %>%
	select(-n, -non_zero_count) %>%
	pivot_wider(names_from=category, values_from = pct) 
}

count_table <- worm_counts_qc_filt_lax
```

```{r, intestine}
marker_genes <- intestine_markers
marker_matrix <- as.matrix(count_table[rownames(marker_genes),])

non_zeros <- get_all_tissues_non0(marker_matrix) %>%
	relocate(intestine, intestine_other, .after=gene)
	
all_tissues <- marker_genes %>%
	tibble::rownames_to_column(var = "gene") %>%
	full_join(non_zeros) %>%
	relocate(gene, pct.1, pct.2, myAUC, power) %>%
	relocate(avg_diff, avg_log2FC, .after=last_col())

cells_of_interest <- filter(all_clusters_categorised, category == "intestine") %>% pull(cell_name)
	
non_zero_days <- get_timepoint_non_zeros(marker_matrix, cells_of_interest) %>%
	filter(day != "d15") %>%
	pivot_wider(names_from = day, values_from = pct, names_prefix = "non0_")

intestine_markers_all <- full_join(all_tissues, non_zero_days)
```

```{r, bw}
marker_genes <- body_wall_muscle_markers
marker_matrix <- as.matrix(count_table[rownames(marker_genes),])

non_zeros <- get_all_tissues_non0(marker_matrix) %>%
	relocate(body_wall, bw_other, .after=gene)
	
all_tissues <- marker_genes %>%
	tibble::rownames_to_column(var = "gene") %>%
	full_join(non_zeros) %>%
	relocate(gene, pct.1, pct.2, myAUC, power) %>%
	relocate(avg_diff, avg_log2FC, .after=last_col())

cells_of_interest <- filter(all_clusters_categorised, category == "body_wall") %>% pull(cell_name)
	
non_zero_days <- get_timepoint_non_zeros(marker_matrix, cells_of_interest) %>%
	filter(day != "d15") %>%
	pivot_wider(names_from = day, values_from = pct, names_prefix = "non0_")

bw_markers_all <- full_join(all_tissues, non_zero_days)
```

```{r, pharynx}
marker_genes <- pharynx_markers
marker_matrix <- as.matrix(count_table[rownames(marker_genes),])

non_zeros <- get_all_tissues_non0(marker_matrix) %>%
	relocate(pharynx, pharynx_other, .after=gene)

all_tissues <- marker_genes %>%
	tibble::rownames_to_column(var = "gene") %>%
	full_join(non_zeros) %>%
	relocate(gene, pct.1, pct.2, myAUC, power) %>%
	relocate(avg_diff, avg_log2FC, .after=last_col())

cells_of_interest <- filter(all_clusters_categorised, category == "pharynx") %>% pull(cell_name)
	
non_zero_days <- get_timepoint_non_zeros(marker_matrix, cells_of_interest) %>%
	filter(day != "d15") %>%
	pivot_wider(names_from = day, values_from = pct, names_prefix = "non0_")

pharynx_markers_all <- full_join(all_tissues, non_zero_days)
```

```{r, neurons}
marker_genes <- neuron_markers
marker_matrix <- as.matrix(count_table[rownames(marker_genes),])

non_zeros <- get_all_tissues_non0(marker_matrix) %>%
	relocate(neurons, other_neurons, .after=gene)

all_tissues <- marker_genes %>%
	tibble::rownames_to_column(var = "gene") %>%
	full_join(non_zeros) %>%
	relocate(gene, pct.1, pct.2, myAUC, power) %>%
	relocate(avg_diff, avg_log2FC, .after=last_col())

cells_of_interest <- filter(all_clusters_categorised, category == "neurons") %>% pull(cell_name)
	
non_zero_days <- get_timepoint_non_zeros(marker_matrix, cells_of_interest) %>%
	filter(day != "d15") %>%
	pivot_wider(names_from = day, values_from = pct, names_prefix = "non0_")

neuron_markers_all <- full_join(all_tissues, non_zero_days)
```
   
In the set of tables below we've now got the percentage of non zero counts for all tissues for each of the marker genes.
The columns named non0_dx (at the far right of the tables) still contain the percentage of cells with non zero counts for the gene and timepoint in the tissue of interest.

# Main results tables

## Intestine 

```{r}
formatDT_no_relocate(intestine_markers_all)
```

## Body wall muscle

```{r}
formatDT_no_relocate(bw_markers_all)
```

## Pharynx

```{r}
formatDT_no_relocate(pharynx_markers_all)
```

## Neurons

```{r}
formatDT_no_relocate(neuron_markers_all)
```


## Table information

The FindMarkers function in Seurat has been run, using the ROC analysis option rather than the default wilcox test. This should be more effective at finding specific markers and also provides more metrics for filtering on. The tables below show the outputs of these tests, along with some additional information for the different timepoints.

Column information:

-   pct.1: The proportion of cells where the gene is detected in the first group. The `first group` includes all of the cells in the clusters specified for that tissue. These are the clusters in green from Harry's Excel file for intestine, body wall muscle and pharynx, and all of the clusters for neurons in the Excel file.

-   pct.2: The proportion of cells where the gene is detected in the second group, where the second group refers to all the remaining cells that are not in the first group.

-   myAUC: Output from the ROC analysis. Info taken from the FindMarkers documentation: "For each gene, evaluates (using AUC) a classifier built on that gene alone, to classify between two groups of cells. An AUC value of 1 means that expression values for this gene alone can perfectly classify the two groupings (i.e. Each of the cells in cells.1 exhibit a higher level than each of the cells in cells.2). An AUC value of 0 also means there is perfect classification, but in the other direction. A value of 0.5 implies that the gene has no predictive power to classify the two groups. Returns a 'predictive power' (abs(AUC-0.5) * 2) ranked matrix of putative differentially expressed genes."  
	We only have AUC values \> 0.5 as we were only testing in one direction.

-   power: predictive power as described above

-   min_day: The minimum value of the following 6 columns - added to enable further filtering

-   non0_d1, non0_d3 etc.: These columns have been added to show information for the individual timepoints. They contain the percentage of cells with non zero counts for the gene and timepoint in the tissue of interest.

-   avg_diff, avg_log2FC: Outputs from the FindMarkers function providing information on differences between group 1 and group 2.

The table is sorted by default by AUC/power, so that the genes at the top should be the best distinguishing markers overall, though they are not necessarily expressed at high levels in all timepoints. The table can be re-sorted by clicking on any column header. Additionally, each of the columns can be filtered by clicking in the box below the column header. This means that minimum/maximum values could be set for any of the columns.

    

# Extended cluster sets {#extended_clusters}

I tried running the ROC analysis including the clusters in white from the Excel sheet. Overall, this didn't seem to help the analysis, but the results from the extended sets can be seen in the tables below.

```{r, eval=first_run}
# process the data once to get the timepoint info
load("data/worm_counts_qc_filt_lax.rds")
load("data/intestine_markers_extended_roc.rds")
load("data/body_wall_muscle_markers_extended_roc.rds")
load("data/pharynx_markers_extended_roc.rds")
load("data/neuron_markers_roc.rds")
load("data/cluster_names.rds")
load("data/worm_meta.rds")

# intestine extended
name_pattern <- "^3_[1245]|^30_[04]|^1_2|^38_0"
of_interest <- grepl(name_pattern, cluster_names) # intestine
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
intestine_table_extended <- timepoint_wrapper(worm_counts_qc_filt_lax, intestine_markers_extended, x)

# body wall muscle extended
name_pattern <- "^4_[012]|^1_1|^7_[01]|^143_0"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
bw_table_extended <- timepoint_wrapper(worm_counts_qc_filt_lax, body_wall_muscle_markers_extended, x)

# pharynx extended
name_pattern <- "^1_0|^26_[012]|^24_1|^41_0|^78_0|^9_[012]|^15_0|^24_0|^41_1"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
x <- rownames(meta[of_interest , ])
pharynx_table_extended <- timepoint_wrapper(worm_counts_qc_filt_lax, pharynx_markers_extended, x)

# neuron hasn't changed

save(intestine_table_extended, file = "data/intestine_table_extended.rds")
save(bw_table_extended, file = "data/bw_table_extended.rds")
save(pharynx_table_extended, file = "data/pharynx_table_extended.rds")
```


```{r}
load("data/intestine_table_extended.rds")
load("data/bw_table_extended.rds")
load("data/pharynx_table_extended.rds")
load("data/neuron_table.rds")
load("data/cluster_names.rds")
```

## Intestine extended

```{r}
formatDT(intestine_table_extended)
```
    
    
## Body wall muscle extended

```{r}
formatDT(bw_table_extended)
```
   
   
## Pharynx extended

```{r}
formatDT(pharynx_table_extended)
```
   

# Checking the cluster names

## Intestine  

```{r}
of_interest <- grepl("^3_[1245]", cluster_names) # intestine
unique(cluster_names[of_interest])

```

## Body wall muscle   

```{r}
name_pattern <- "^4_[012]|^1_1"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
```

## Pharynx  

```{r}
name_pattern <- "^26_[012]|^24_1|^41_0|^78_0"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
```

## Neurons  

```{r}
name_pattern <- "^6_[0123]|^10_0|^14_[012]|^1[79]_0|^20_[012]|^22_0|^25_[01]|^2[89]_0|^32_0|^35_[01234]|^36_[023]|^39_[0123]|^44_[01]|^46_0|^47_[01]|^4[89]_0|^5[01245]_0|^51_1|^6[1235689]_0|^70_[01]|^7[1234567]_0|^8[0134567]_0|^9[01234789]_0|^10[01346789]_0|^11[1234578]_0|^12[0123789]_0|^13[012345678]_0|^14[0245]_0"
of_interest <- grepl(name_pattern, cluster_names)

no <- tibble::tibble(cluster_name = unique(cluster_names[of_interest])) %>%
	tidyr::separate_wider_delim(cluster_name, delim = ":", names=c("cluster_no", "cluster_name")) %>%
	pull(cluster_no)

#unique(cluster_names[of_interest])

tibble::tibble(cluster_name = unique(cluster_names[of_interest])) %>%
	tidyr::separate_wider_delim(cluster_name, delim = ":", names=c("cluster_no", "cluster_name")) %>%
	tidyr::separate_wider_delim(cluster_no, delim="_", names=c("first_no", "second_no")) %>%
	dplyr::mutate(first_no = as.numeric(first_no)) %>%
	arrange(first_no, second_no) %>%
	tidyr::unite(col="numbers", first_no, second_no, sep="_") %>%
	tidyr::unite(col="cluster", numbers, cluster_name, sep=":") %>%
	pull(cluster)
# DT::datatable(rownames = FALSE)


# checking against the set from Excel	
#neuron_clusters <- paste0("cluster_", no)
	
```

### Intestine_extended

```{r}
name_pattern <- "^3_[1245]|^30_[04]|^1_2|^38_0"
of_interest <- grepl(name_pattern, cluster_names) # intestine
unique(cluster_names[of_interest])
```

### Body wall muscle extended

```{r}
name_pattern <- "^4_[012]|^1_1|^7_[01]|^143_0"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
```

### Pharynx extended

```{r}
name_pattern <- "^1_0|^26_[012]|^24_1|^41_0|^78_0|^9_[012]|^15_0|^24_0|^41_1"
of_interest <- grepl(name_pattern, cluster_names)
unique(cluster_names[of_interest])
```








